The Concurrent Calculi Formalisation Benchmark

### 2024-02-22 short-paper/short-paper.tex
merge
Fix some minor comments.

\begin{abstract}
  POPLMark and POPLMark Reloaded sparked a flurry of work on machine-checked proofs, and fostered the adoption of proof mechanisation in programming language research.
  Both challenges were purposely limited in scope, and their benchmark problems do not address concurrency-related issues.
  %
  We propose a new collection of benchmark problems focused on the difficulties that typically arise when mechanising formal models of concurrent and distributed programming languages, such as process calculi.
  Our benchmark problems address three key topics: linearity, scope extrusion, and coinductive reasoning.
  The goal of this new challenge is to clarify, compare, and advance the state of the art, fostering the adoption of proof mechanisation in future research on message-passing concurrency. \todo{Broaden `message-passing concurrency' to `concurrency'?}

@@

In this spirit, we introduce a new collection of benchmark problems
specifically crafted to tackle common challenges encountered during
the mechanisation of formal models for concurrent and distributed
programming languages.
We particularly focus on process calculi.\todo{Why? - they are PL-like but also formal models in themselves}
Concurrent calculi are notably subtle: for instance, it took nine
years before the incorrect proof of type safety in the seminal paper
by Honda et al.~\cite{Honda1998} was rectified in Yoshida and
Vasconcelos’s subsequent work\cite{Yoshida2007}.
Similarly, several key results in papers on session types have turned out
to be false later on, illustrating the
value of machine-checked proofs~\cite{Gay2020,10.1145/3290343}.

@@

Our benchmark problems (detailed in \cref{app:challenges} and on our website) consider
\emph{in isolation} three key aspects that frequently pose challenges
when mechanising concurrency theory: \emph{linearity}, \emph{scope
  extrusion}, and \emph{coinductive reasoning}.  Mechanisations must
often address several of these aspects at the same time; however, we
see the combination of techniques as a next
step, as discussed in \cref{sec:going-beyond}.
Our benchmark challenges are based on process calculi, as these provide a simple but
realistic showcase of common issues encountered during mechanisation
efforts.

@@

Given those peculiarities, the syntax and semantics of $\pi$-calculi have been
mechanised from an early age (see~\cite{Melham1994}) with many proof
assistants and in so many encoding styles that it would take too long to
review.  They range from concrete encodings based on named syntax
\cite{Melham1994} to basic de Bruijn \todo{cite} and locally nameless
representation %~\cite{Chargueraud2012}
(\eg~\cite{Castro2020}).
Nominal approaches are also common (see~\cite{Bengtson2009}), but they
may be problematic in proof assistants based on dependent types.

@@

When reasoning over bisimilarity many authors rely on the native
 coinduction offered by the chosen proof assistants
\cite{Bengtson2016,Kahsai2008,Thiemann2019,Gay2020}, while others prefer a
more ``set-theoretic'' approach
\cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}. Some
use both and establish an internal adequacy~\cite{Honsell2001}. Few
extend the proof assistant foundations to allow \eg reasoning about
bisimilarity up-to~\cite{ChaudhuriCM15}.

>>>>

\begin{abstract}
  POPLMark and POPLMark Reloaded sparked a flurry of work on machine-checked proofs, and fostered the adoption of proof mechanisation in programming language research.
  Both challenges were purposely limited in scope, and their benchmark problems do not address concurrency-related issues.
  %
  We propose a new collection of benchmark problems focused on the difficulties that typically arise when mechanising formal models of concurrent and distributed programming languages, such as process calculi.
  Our benchmark problems address three key topics: linearity, scope extrusion, and coinductive reasoning.
  The goal of this new challenge is to clarify, compare, and advance the state of the art, fostering the adoption of proof mechanisation in future research on concurrency.

@@

In this spirit, we introduce a new collection of benchmark problems
specifically crafted to tackle common challenges encountered during
the mechanisation of formal models for concurrent and distributed
programming languages.
We particularly focus on process calculi as these provide a simple but
realistic showcase of common issues encountered during mechanisation
efforts.
Concurrent calculi are notably subtle: for instance, it took nine
years before the incorrect proof of type safety in the seminal paper
by Honda et al.~\cite{Honda1998} was rectified in Yoshida and
Vasconcelos’s subsequent work\cite{Yoshida2007}.
Similarly, several key results in papers on session types have turned out
to be false later on, illustrating the
value of machine-checked proofs~\cite{Gay2020,10.1145/3290343}.

@@

Our benchmark problems (detailed in \cref{app:challenges} and on our website) consider
\emph{in isolation} three key aspects that frequently pose challenges
when mechanising concurrency theory: \emph{linearity}, \emph{scope
  extrusion}, and \emph{coinductive reasoning}.  Mechanisations must
often address several of these aspects at the same time; however, we
see the combination of techniques as a next
step, as discussed in \cref{sec:going-beyond}.

@@

Given those peculiarities, the syntax and semantics of $\pi$-calculi have been
mechanised from an early age (see~\cite{Melham1994}) with many proof
assistants and in so many encoding styles that it would take too long to
review.  They range from concrete encodings based on named syntax
\cite{Melham1994} to basic de Bruijn~\cite{Hirschkoff1997, Perera2018} and locally nameless
representation %~\cite{Chargueraud2012}
(\eg~\cite{Castro2020}).
Nominal approaches are also common (see~\cite{Bengtson2009}), but they
may be problematic in proof assistants based on dependent types.

@@

When reasoning over bisimilarity many authors rely on the native
 coinduction offered by the chosen proof assistants
\cite{Bengtson2016,Kahsai2008,Thiemann2019,Gay2020}, while others prefer a
more ``set-theoretic'' approach
\cite{Hirschkoff1997,Bengtson2009,Maksimovic2015,Pohjola2022}. Some
use both and establish an internal adequacy~\cite{Honsell2001}. Few
extend the proof assistant foundations to allow \eg reasoning about
bisimilarity up-to~\cite{ChaudhuriCM15}.
% While many proof assistants support
% coinductive techniques,   but they support subtly different formalisms that enable different techniques, and some systems provide more than one abstraction to use coinduction.
% We thus see a variety of different formalisation choices for
% bisimulations, sometimes ad-hoc for the specific result.

% \todo{Reorganize the rest of this section to be more coherent}
% Isabelle, Coq, and Abella seamlessly support coinductive definitions and reasoning.
% Agda and Beluga support it through the convenient abstraction of co-patterns~\cite{Abel2013}.
% Newer entrants to the field like the Lean proof assistant
% are working on adding a notion of coinduction to their system~\cite{Avigad2019}.\alb{I kinda object to ``seamlessly''. In Coq it's a mess, in Abella very very limited, sized types in Agda are shaky. I guess Isabelle is better}
% =======
% Process calculi often include constructs that allow processes to have infinite behaviours.
% Coinduction is a fundamental technique for defining and reasoning about infinite objects, and it can thus be used to reason about these behaviours.
% While many proof assistants support coinductive techniques, using them is notoriously difficult, and representations and proof techniques vary widely.
% The crux of our challenge is the effective use of coinductive up-to techniques.
% The intention is that it should be easy to use the bisimulation relation and its properties, but this may be difficult with some mechanisation approaches.

% The goal of our challenge on coinductive reasoning is proving that strong barbed bisimilarity can be turned into a congruence by making it sensitive to substitution and parallel composition.
% The setting for our challenge is an untyped \(\pi\)-calculus with replication of processes enabling infinite behaviours.
% We do not include delegation in the calculus since it is orthogonal to our primary aim of exploring coinductive proof techniques.
% We base our definition of bisimilarity on a labelled transition system semantics and an observability predicate describing the communication steps available to a process.
% The introduction of strong barbed bisimulation is one of the first steps when studying the observable behaviour of process calculi, both in classical textbooks (\eg, \cite{picalcbook}) and in existing mechanisations.
% Though weak barbed congruence is a more common behavioural equivalence, we prefer strong equivalences to simplify the theory by avoiding the need to abstract over the number of internal transitions in a trace.

% Coinduction is widely supported in proof assistants, but they support subtly different formalisms that enable different techniques, and some systems provide more than one abstraction to use coinduction.
% We thus see a variety of different formalisation choices for bisimulations, sometimes ad-hoc for the specific result.

% \todo{Reorganize the rest of this section to be more coherent}
% Isabelle, Coq, and Abella seamlessly support coinductive definitions and reasoning.
% Agda and Beluga support it through the convenient abstraction of co-patterns~\cite{Abel2013}.
% Newer entrants to the field like the Lean proof assistant
% are working on adding a notion of coinduction to their system~\cite{Avigad2019}.\alb{I kinda object to ``seamlessly''. In Coq it's a mess, in Abella very very limited, sized types in Agda are shaky. I guess Isabelle is better}
